\section{The Data}
\label{sec:data}

This study focuses on stellar rotation in the original \kepler\ field, partly
because \kepler\ provides the largest samples of homogeneously measured
rotation periods, and partly because its low Galactic latitude allows us to
marginalize over missing RV measurements and precisely infer vertical
velocity, \vz.
We combined two large rotation period catalogs constructed from original
\kepler\ data: \mct\ and \sant.
These two studies used different techniques to measure rotation periods from
\kepler\ light curves: autocorrelation functions and wavelets respectively.
The \citet{santos2019} study was specifically focused on cooler stars: K and M
dwarfs, and includes a larger number of rotation periods for these stars.
The combined catalogs provided rotation periods for a total of 38,710 stars.

We used the publicly available \kepler-Gaia DR2 crossmatched
catalog\footnote{Available at gaia-kepler.fun} to combine the \mct\ and \sant\
rotation catalogs with the Gaia DR2 catalog of parallaxes, proper motions
and apparent magnitudes.
Reddening and extinction from dust was calculated for each star using the
Bayestar dust map implemented in the {\tt dustmaps} {\it Python} package
\citep{green2018}, and {\tt astropy} \citep{astropy2013, astropy2018}.
We used Gaia DR2 photometric color, $G_{\rm BP} - G_{\rm RP}$, to estimate
effective temperatures for the stars in our sample, using the calibrated
relation in \citet{curtis2020}.

% Unlike isolated main-sequence stars, the rotation periods of binary stars and
% subgiants cannot always be determined by their mass and age (or at least they
% do not always follow the {\it same} gyrochronology relationship as isolated
% dwarfs).
% Photometric binaries and subgiants were therefore removed from the sample by
% applying cuts to the color-magnitude diagram (CMD), shown in figure
% \ref{fig:CMD}.
% A 6th-order polynomial was fit to the main sequence and raised by 0.27 dex to
% approximate the division between single stars and photometric binaries (shown
% as the curved dashed line in figure \ref{fig:CMD}).
% All stars above this line were removed from the sample.
% Potential subgiants were also removed by eliminating stars brighter than 4th
% absolute magnitude in Gaia G-band.
% This cut also removed a number of main sequence F stars from our sample,
% however these hot stars are not the focus of our gyrochronology study since
% their small convective zones inhibit the generation of a strong magnetic
% field.
% The removal of photometric binaries and evolved/hot stars reduced the total
% sample of around 38,000 stars by around 4,000.

3705 stars in our sample had RV measurements available in Gaia DR2, with a
median uncertainty of 1.88 \kms.
Gaia DR2 included RVs for stars with Gaia apparent magnitudes between 4
and 13, and 3550 K $\lesssim$ \teff\ $\lesssim$ 6900 K \citep{brown2018}.
We also crossmatched the \mct\ sample with the 5th LAMOST data release
\citep{cui2012, xiang2019}, adding a further 10623 RV measurements to the
sample, and expanding the total number of stars with measured RVs to 14,328.
The median uncertainty of the LAMOST RV measurements was 4.71 \kms\ and,
given that the Gaia RVs were more precise, on average, than the LAMOST
RVs, we adopted the Gaia value in cases where both were available.
We note that the third Gaia data release will contain a large number of new
RV measurements for the stars in our sample.

We removed stars with a Gaia parallax signal-to-noise ratio of less than 10,
stars with negative parallaxes, and stars with a Gaia astrometric excess noise
value greater than 5.
After these cuts, 35,328 stars remained in our sample, of which 11,050 had RV
measurements from either Gaia or LAMOST.
We calculated 3D velocities for all stars with measured RVs using {\tt
astropy}, and {\it inferred} 3D velocities for the remaining 24,278 stars
using the method described in section \ref{sec:velocities}.
